Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-03 paper/paper.tex
Rewrote some chunks of 3.1

\Rref{I-Pi} is the most complex rule in our logical relation.
Consider first the following simplified \rref{I-PiAlt}, which follows from \rref{I-Pi}.

@@

\Rref{I-PiAlt} directly captures the meaning of
a well-behaved dependent function type. The preconditions of
the rule says that the function type $[[Pi x : A . B]]$ has an interpretation
if its input type $[[A]]$ can be interpreted as some set $[[S]]$, and for all
terms $[[a in S]]$, the type $[[B {a / x}]]$, obtained by substituting $[[a]]$
into the output type $[[B]]$, has some semantic interpretation $[[S0]]$. In its
conclusion, the  interpretation of $[[Pi x : A . B]]$ is the set of terms
$[[b]]$ such that for all $[[a in S]]$, where $[[S]]$ is an interpretation of
$[[A]]$, the application form $[[b a]]$ belongs to all possible interpretations
of $[[B {a  / x}]]$. The second precondition ensures at least one interpretation
exists for each $[[B {a / x}]]$ where $[[a in S]]$.

\jc{I think at this point it's unclear why \rref{I-PiAlt} is shaped the way it is.
I found myself asking: If in the second precondition we've established the existence of $[[S0]]$,
why do we need to quantify over all interpretations $[[S0]]$?
In fact, instead of the second precondition,
why can't split it into two and say that there exists some family of sets
$S_0 : \forall a, a \in S \to [[PowerSet STm]]$
such that $\forall a, (m : a \in S) \to [[Interp I i B { a / x } (# S0 a m #)]]$,
then conclude that the intepretation of the function type yields a set over $[[b]]$
such that $\forall a, (m : a \in S) \to [[b in S0 a m]]$?
As far as I can tell, the interpretation would be well formed,
but proving the fundamental theorem would then require something like indefinite description,
since the induction hypothesis would yield $\forall a, a \in S \to \exists S_0, \dots$,
the mere existence of such a family of sets,
while the constructor requires the actual family of sets itself.
This is subtle but maybe warrants explaining.}

\begin{lemma}[I-PiAlt derivability]
  \label{lemma:piintroalt}
  \Rref{I-PiAlt} is derivable from \rref{I-Pi}.
\end{lemma}
\begin{proof}
The precondition $[[forall a, (# a in S
implies (# exists S0 , Interp I i B { a / x } S0 #) #)]]$ from
\rref{I-PiAlt} immediately induces a function $[[F in S -> PowerSet
STm]]$ such that $[[forall a, (# a in S implies Interp I i B { a / x }
F a #)]]$, which is exactly what we need to apply \rref{I-Pi}.
\end{proof} 
In fact, while \rref{I-PiAlt} is an instantiation of \rref{I-Pi},
these two rules are equivalent in the sense that every derivation
involving \rref{I-Pi} can be systematically replaced by
\rref{I-PiAlt}. This equivalence follows directly from the fact
that the logical relation is a partial function, a result we will show
in Lemma~\ref{lemma:logreldeter}. The preconditions of \rref{I-Pi},
when combined with the functionality of the logical relation, uniquely
determine the function $[[F in S -> PowerSet STm]]$ to be the
functional relation
$[[{ (a , S0 ) | a in S implies Interp I  i B { a /x } S0 }]]$. This result is
formally shown through the improved inversion lemma for function types
(Lemma~\ref{lemma:piinvalt}).

Unfortunately, we cannot define the function case of our logical
relation directly using \rref{I-PiAlt} since the occurrence of
$[[Interp I i B {a/x} S0]]$ in its conclusion not only violates the
syntactic strict positivity constraint required in proof assistants,
but is genuinely non-monotone when we treat the inductive definition
as the fixed point of an endofunction over the domain of relations.
Intuitively, the failure of monotonicity stems from the fact
that the witness picked in the precondition is not necessarily the
same witness being referred to in the post condition as the relation
grows, whereas the
function $[[F]]$ in \rref{I-Pi} ``fixes'' the witnesses $[[S0]]$ as
$[[F a]]$ for each $[[a in S]]$,
thus preventing the set of witnesses from growing. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties to justify the well-definedness of our
inductive relation with \rref{I-PiAlt}, we opt for our current
\rref{I-Pi} that immediately produces a
well-defined inductive relation and usable induction principle. The
slight disadvantage of \rref{I-Pi} is that we need to construct the
function $[[F]]$ each time we apply it, though this is mitigated by
the derivability of \rref{I-PiAlt} and the alternative $\Pi$ inversion
principle (Lemma~\ref{lemma:piinvalt}).

>>>>

\Rref{I-Pi} is the most complex rule in our logical relation.
It states that, to build the interpretation of a dependent function type,
we first require the interpretation $[[S]]$ of its domain $[[A]]$.
Because the codomain depends on the function input,
rather than a single interpretation, we require an interpretation $[[F a]]$
of the codomain $[[B {a / x}]]$ for each $[[a in S]]$.
The interpretation of the overall function type is the set of terms $[[b]]$
such that for every term $[[a]]$ in the interpretation $[[S]]$ of $[[A]]$,
$[[b a]]$ is in the interpretation $[[F a]]$ of $[[B {a / x}]]$.

However, this form of the rule is less convenient to work with, since it
requires constructing the sets $[[F]]$ separately from the proof that they are
indeed interpretations of $[[B]]$. A more convenient form is the following
\rref{I-PiAlt}, which combines the existence of interpretations of the codomain
with their proofs.

@@

The second precondition now states that for every $[[a in S]]$,
there must exist some interpretation $[[S0]]$ of $[[B {a / x}]]$.
The interpretation of the overall function type then states that $[[b a]]$
must be in every interpretation $[[S0]]$ of $[[B {a / x}]]$.
Later, Lemma~\ref{lemma:logreldeter} proves functionality of the interpretation
of types, which guarantees that there must only be one such interpretation.
This alternate rule for the interpretation of function types follows from \rref{I-Pi}.

\begin{lemma}[I-PiAlt derivability\footnote{\dotv{semtyping.v}{InterpExt\_Fun\_nopf}}]
  \label{lemma:piintroalt}
  \Rref{I-PiAlt} is derivable from \rref{I-Pi}.
\end{lemma}

\begin{proof}
  The precondition
  $[[forall a, (# a in S implies (# exists S0 , Interp I i B { a / x } S0 #) #)]]$
  from \rref{I-PiAlt} immediately induces a function
  $[[F in S -> PowerSet STm]]$ such that
  $[[forall a, (# a in S implies Interp I i B { a / x } F a #)]]$,
  which is exactly what we need to apply \rref{I-Pi}.
\end{proof}

While \rref{I-PiAlt} is an instantiation of \rref{I-Pi}, by functionality,
the two rules are equivalent in the sense that every derivation involving
\rref{I-Pi} can be systematically replaced by \rref{I-PiAlt}.
Furthermore, functionality uniquely determines the function
$[[F in S -> PowerSet STm]]$ to be the functional relation
$[[{ (a , S0 ) | a in S implies Interp I i B { a /x } S0 }]]$.
This result is shown by Lemma~\ref{lemma:piinvalt},
the corresponding inversion lemma for \rref{I-PiAlt}.

Unfortunately, we cannot directly define the interpretation of function types
by \rref{I-PiAlt}, since the occurrence of $[[Interp I i B {a/x} S0]]$ in its
conclusion not only violates the syntactic strict positivity constraint on
inductive definitions required by proof assistants,
but is genuinely non-monotone when we treat the inductive definition
as the fixed point of an endofunction over the domain of relations.
\jc{What?}
Intuitively, the failure of monotonicity stems from the fact that the witness
picked in the precondition is not necessarily the same witness being referred
to in the postcondition as the relation grows, \jc{Why?} whereas the function $[[F]]$
in \rref{I-Pi} ``fixes'' the witnesses $[[S0]]$ as $[[F a]]$ for each
$[[a in S]]$, thus preventing the set of witnesses from growing. While it might
be possible to restrict the domain with additional constraints such as
functionality and inversion properties to justify the well-definedness of our
inductive relation with \rref{I-PiAlt}, we opt for our current
\rref{I-Pi} that immediately produces a
well-defined inductive relation and usable induction principle.
Then by deriving \rref{I-PiAlt} and its inversion lemma,
we avoid needing to manipulate the function $[[F]]$ directly.
